|

### 2024-08-21 reproducibility/manuscript/manuscript.qmd
docs(manuscript): use v2 cached Fig. 1 and Ext Fig. 1
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
fix(manuscript): use bold panel references
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@

We next used Pyro&thinsp;-Velocity to unbiasedly uncover potential cell fate
determinant markers genes within the E15.5 mouse pancreas. To this end, we first
selected the top 300 genes with the best goodness of fit (negative mean absolute
errors across cells). Next, we ranked them based on the positive (or negative)
correlation between inferred shared time and corresponding denoised spliced
expression per cell ([@fig-model-results]d, Methods). This procedure
successfully recovered key developmental genes that are activated (or repressed)
during endocrinogenesis ([@fig-model-results]d-e, @extfig-model-results). Among
the top positively correlated genes identified by Pyro&thinsp;-Velocity,
endocrine Beta cells expressed insulin 1 and 2 (Ins1, Ins2) and Nnat, which is a
key regulator Beta cell insulin generation [@Millership2018-vd] and Ppp1r1a
which is a marker of Beta cells [@Jiang2013-pt]. The ranking procedure proposed
by scVelo failed to identify Ins1, Ins2, Nnat, and Ppp1r1a as lineage-restricted
genes in their top 50 ranked gene list. The inability of scVelo to identify
these key developmental genes could be likely explained by the kNN smoothing of
read count and filtering steps (Supplementary Table 1, 2). Visual inspection of
these four top genes by Pyro&thinsp;-Velocity revealed a single activation phase
during cell type transitions based on phase portrait curves. We also propose
their visualization through ``rainbow plots'' to illustrate denoised single gene
spliced expression across shared time and to investigate the inferred cell order
based on known annotations. Further, UMAP embedding of cells colored based on
denoised gene expression confirmed the cell fate markers' smooth and progressive
activation (or repression) towards endocrine cells ([@fig-model-results]e,
[@extfig-model-results]).

@@

::: {#fig-model}
![](figures/fig.pdf)

Pyro&thinsp;-Velocity is a fully generative Bayesian method with uncertainty
estimation of velocity vector fields and shared latent time, based on raw counts
and without ad-hoc preprocessing steps. a. Schematic comparing current RNA
Velocity and Pyro&thinsp;-Velocity workflows. Both methods take in input spliced
and unspliced read count tables and output RNA velocity vector fields and latent
(shared) time. Previous methods require ad-hoc preprocessing steps, including
the arbitrary selection of top variable genes, number of principal components,
and kNN-pooling to impute both spliced and unspliced expression within the
PCA-reduced spliced expression space. In addition, manually tuned ad-hoc
postprocessing steps are used to compute the final vector field and latent time
based on aggregation and smoothing of the independent velocity estimates for
each gene which can dramatically change the velocity estimates. In contrast,
Pyro&thinsp;-Velocity is a fully generative and probabilistic model that uses
raw read counts. Our model requires minimal preprocessing and postprocessing
steps and outputs a common shared time based on all the genes without additional
aggregation steps. It also provides uncertainty estimation based on posterior
samples. b. The main outputs of the Pyro&thinsp;-Velocity framework.
Visualization of single gene expression across inferred developmental cell order
using rainbow plots (right panel).
:::

@@

Pyro&thinsp;-Velocity outperforms other RNA-velocity methods and accurately
predicts cell lineage trajectories. a,b. RNA Velocity model comparisons from
fully mature peripheral blood mononuclear cells (PBMCs, t-SNE embedded plots, a)
or E15.5 mouse pancreas (UMAP embedded plots, b-c).  From left to right in
panels a and b: 1. Cells colored by cell type; 2. Stream plot of velocity vector
field defined by scVelo; 3. Stream plot of the posterior mean vector field from
Pyro&thinsp;-Velocity; 4. Pyro&thinsp;-Velocity visualization of cell dynamics
for 3 arbitrarily selected single cells; 5. Single-cell vector field uncertainty
where each cell is colored based on the scaled angular standard deviation to
visualize uncertainty based on n=30 iterated, posterior samples depicted as
arrows noted throughout our analysis; 6. Averaged vector field uncertainty where
each vector is colored by the averaged vector field uncertainty. The scaled
angular standard deviation range is [0 360], where 0 corresponds to the highest
confidence for a single direction, 360 corresponds to the highest uncertainty,
and no directionality preference based on 30 posterior samples (4-6). Vectors
were projected into the UMAP space using the scVelo API based on cosine
similarity-derived transition matrix. c. UMAP embedded cells colored by the
average cell shared time across 30 posterior samples. The Spearman's correlation
between Pyro&thinsp;-Velocity and Cytotrace is provided. d.
Pyro&thinsp;-Velocity automatic gene selection and ranking for predicting
important genes involved in pancreas development. e. Phase portraits of
posterior samples average spliced and unspliced gene expression (left), rainbow
plots showing denoised spliced gene expression (middle), and gene expression
within UMAP embedded space.  Single dots correspond to individual cells (left,
right) or vertical lines (middle). Cell type annotation coloring is the same as
in panel b.
:::

@@

Pyro&thinsp;-Velocity accurately predicts cell fate choices and lineage
trajectory in human LARRY barcoded hematopoietic cells. Systematic benchmarking
of cell fate prediction methods - predicting uni-lineage cell fates (a-b, Model
1), bifurcated lineages (c, Model 2), or all lineages together (d, Model 2).
Panels from left to right show on the same SPRING embedding: 1. Cell types; 2.
Clonal progression vector field derived based on the observed clonal barcodes
(Methods); 3. scVelo recovered vector field; 4. Pyro&thinsp;-Velocity single
cell level vector field uncertainty (angular standard deviation from 30
posterior samples);
5. Pyro&thinsp;-Velocity recovered vector field; vectors are colored by the
averaged vector field uncertainty; 6. Clonal fate potency score using Cospar
[@Wang2022-xb]; 7. scVelo recovered latent time; 8. Pyro&thinsp;-Velocity
inference of posterior mean shared time per cell. Cosine similarity and
Spearman's correlation for the different methods with the clonal progression and
fate potency are reported in the titles.
:::

@@

::: {#extfig-graphical-models}
![](figures/fig.pdf)

Graphical representations detail the generative process of the
Pyro&thinsp;-Velocity models and highlight their differences. a.
Pyro&thinsp;-Velocity Model 1. Each rectangle represents one data dimension
(e.g., gene or cell). The arrows represent the directed hierarchical dependency
among random variables. The white circles correspond to the modeled random
variables. The probability distributions of these variables are denoted on the
right legend. The gray circles correspond to intermediate results computed from
random variables. In this model, the kinetics random variables for the cell
shared time and gene are modeled with two constraints: 1. the shared time is
strictly larger than the gene-specific activation time $t_{0}^{(0)}$, 2. the
gene expression in the activate state starts with the initial condition
$(u_{0}=0,\,s_{0}=0)$, as in previous methods[@La_Manno2018-lj;@Bergen2020-pj]
(Methods). b. Pyro&thinsp;-Velocity Model 2. This figure follows the same
graphical convention as in a. In this model $t_{0}^{(0)}$ is independent from
the shared time, thus allowing a time lag for the gene activation phase. In
addition, a positive starting basal gene expression level $(u_{0},\,s_{0})$
different than $(u_{0}=0,\,s_{0}=0)$ is allowed. (Methods). c. Schematic
illustrating how the constraints proposed in Model 1 (left) and Model 2 (right)
are reflected in the modeling of individual gene expression. The x-axis
corresponds to the shared time, the y-axis corresponds to the posterior mean of
the spliced expression level, the vertical lines with arrows show the start and
end time of the observed cell sampling, t0 represents the gene-specific
activation time $t_{0}^{(0)}$, the red lines highlight the differences for the
potential gene-specific time lag when $t_{0}^{(0)} > {\min}\!⁡(t)$, and the
initial spliced gene expression. The same differences can also be illustrated
for the unspliced gene expression profiles (not shown).
:::

@@

Extended version of @fig-model-results for evaluating Pyro&thinsp;-Velocity
(Model 1) and scVelo on the PBMC and pancreas datasets. a. Left figure shows the
scVelo vector field using the default user options, right figure shows the
evaluation of the scVelo vector field using different combinations of user
options within Epsilon cells. b. Pyro&thinsp;-Velocity additional single cell
level uncertainty diagnostic plots on the fully mature PBMC dataset, from left
to right: 1. standard deviation of posterior samples of shared time, 2. Rayleigh
test of posterior samples of the vector field (the title reports false discovery
rate after p-value correction for multiple tests across cells), 3.
transcriptional state uncertainty of predicted raw read count per cell. The
contour lines show the top 10% of the most uncertain cells. c. For the pancreas
dataset, the figures show from left to right: UMAP rendering of 1. Cytotrace, 2.
scVelo latent time, 3. Posterior mean of Pyro&thinsp;-Velocity shared time
averaged across 30 posterior samples, 4. Top 50 markers overlapping between
Pyro&thinsp;-Velocity and scVelo (Methods, Supplementary Table 1-2), 5. The same
uncertainty diagnostic plots per cell as in b. d. The same plot as Fig. 2d. e.
Selection of the top negatively correlated genes with the cell shared time based
on the posterior mean on denoised spliced expression and negative mean absolute
error of gene expression prediction (Methods). f. Phase portraits, rainbow
plots, and UMAP rendering of posterior mean of gene expression for the genes
selected in e.
:::

@@

Extension of @fig-lineage-tracing, Pyro&thinsp;-Velocity (Model 1 for uni-fate
data, Model 2 for bi-fate and multi-fate data) outperforms Cytotrace on
scRNA-seq datasets with lineage barcoding information (LARRY dataset). a. SPRING
embedding rendering of Cytotrace scores, the title reports the Spearman's
correlation between the Cytotrace and Cospar fate potency scores, b.
Pyro&thinsp;-Velocity additional single cell level uncertainty diagnostic plots
based on the multi-fate LARRY dataset, from left to right: 1. Standard deviation
of posterior samples of shared time per cell, 2. Rayleigh test of posterior
samples of vector field per cell (the title reports the false discovery rate
after p-value correction for multiple tests across cells), 3. Transcriptional
state uncertainty of predicted raw read count per cell. The contour lines show
the top 10% of the most uncertain cells (Methods), c. Pairwise Spearman's
correlation between Cytotrace/Cospar fate potency scores and velocity methods
for the four different benchmarking datasets selected based on the lineage
barcoding information from LARRY data. d. SPRING embedding rendering of
posterior mean of shared time across 30 posterior samples, the title shows the
Spearman's correlation of mean shared time with Cytotrace. e. Selection of the
top positively correlated genes with the cell shared time based on the posterior
mean of spliced expression. f. Phase portraits, rainbow plots, and UMAP
rendering with splicing gene expression level for the selected genes in e.
:::

@@

Extension of [@fig-lineage-tracing] presenting an extended comparison of
Pyro&thinsp;-Velocity and scVelo on scRNA-seq datasets with lineage barcoding
information (LARRY dataset). a. uni-fate Monocyte lineage analysis with
Pyro&thinsp;-Velocity Model 2, b. uni-fate Neutrophil lineage analysis with
Pyro&thinsp;-Velocity Model 2, c. Neutrophil and Monocyte bi-fate lineage
analysis with Pyro&thinsp;-Velocity Model 1, and d. multi-fate lineage analysis
with Pyro&thinsp;-Velocity Model 1. The panel descriptions are the same as in
[@fig-lineage-tracing].
:::

>>>>

    citecolor: NavyBlue
    linkcolor: NavyBlue
    toccolor: NavyBlue
    urlcolor: NavyBlue

@@

We next used Pyro&thinsp;-Velocity to unbiasedly uncover potential cell fate
determinant markers genes within the E15.5 mouse pancreas. To this end, we first
selected the top 300 genes with the best goodness of fit (negative mean absolute
errors across cells). Next, we ranked them based on the positive (or negative)
correlation between inferred shared time and corresponding denoised spliced
expression per cell ([@fig-model-results]d, [Methods](#sec-methods)). This procedure
successfully recovered key developmental genes that are activated (or repressed)
during endocrinogenesis ([@fig-model-results]d-e, @extfig-model-results). Among
the top positively correlated genes identified by Pyro&thinsp;-Velocity,
endocrine Beta cells expressed insulin 1 and 2 (Ins1, Ins2) and Nnat, which is a
key regulator Beta cell insulin generation [@Millership2018-vd] and Ppp1r1a
which is a marker of Beta cells [@Jiang2013-pt]. The ranking procedure proposed
by scVelo failed to identify Ins1, Ins2, Nnat, and Ppp1r1a as lineage-restricted
genes in their top 50 ranked gene list. The inability of scVelo to identify
these key developmental genes could be likely explained by the kNN smoothing of
read count and filtering steps (Supplementary Table 1, 2). Visual inspection of
these four top genes by Pyro&thinsp;-Velocity revealed a single activation phase
during cell type transitions based on phase portrait curves. We also propose
their visualization through ``rainbow plots'' to illustrate denoised single gene
spliced expression across shared time and to investigate the inferred cell order
based on known annotations. Further, UMAP embedding of cells colored based on
denoised gene expression confirmed the cell fate markers' smooth and progressive
activation (or repression) towards endocrine cells ([@fig-model-results]e,
[@extfig-model-results]).

@@

::: {#fig-model}
![](figures/v2/Figure1.pdf)

Pyro&thinsp;-Velocity is a fully generative Bayesian method with uncertainty
estimation of velocity vector fields and shared latent time, based on raw counts
and without ad-hoc preprocessing steps. **a**, Schematic comparing current RNA
Velocity and Pyro&thinsp;-Velocity workflows. Both methods take in input spliced
and unspliced read count tables and output RNA velocity vector fields and latent
(shared) time. Previous methods require ad-hoc preprocessing steps, including
the arbitrary selection of top variable genes, number of principal components,
and kNN-pooling to impute both spliced and unspliced expression within the
PCA-reduced spliced expression space. In addition, manually tuned ad-hoc
postprocessing steps are used to compute the final vector field and latent time
based on aggregation and smoothing of the independent velocity estimates for
each gene which can dramatically change the velocity estimates. In contrast,
Pyro&thinsp;-Velocity is a fully generative and probabilistic model that uses
raw read counts. Our model requires minimal preprocessing and postprocessing
steps and outputs a common shared time based on all the genes without additional
aggregation steps. It also provides uncertainty estimation based on posterior
samples. **b**, The main outputs of the Pyro&thinsp;-Velocity framework.
Visualization of single gene expression across inferred developmental cell order
using rainbow plots (right panel).
:::

@@

Pyro&thinsp;-Velocity outperforms other RNA-velocity methods and accurately
predicts cell lineage trajectories. **a,b**, RNA Velocity model comparisons from
fully mature peripheral blood mononuclear cells (PBMCs, t-SNE embedded plots, a)
or E15.5 mouse pancreas (UMAP embedded plots, **b-c**).  From left to right in
panels **a** and **b**: 1. Cells colored by cell type; 2. Stream plot of velocity vector
field defined by scVelo; 3. Stream plot of the posterior mean vector field from
Pyro&thinsp;-Velocity; 4. Pyro&thinsp;-Velocity visualization of cell dynamics
for 3 arbitrarily selected single cells; 5. Single-cell vector field uncertainty
where each cell is colored based on the scaled angular standard deviation to
visualize uncertainty based on n=30 iterated, posterior samples depicted as
arrows noted throughout our analysis; 6. Averaged vector field uncertainty where
each vector is colored by the averaged vector field uncertainty. The scaled
angular standard deviation range is [0 360], where 0 corresponds to the highest
confidence for a single direction, 360 corresponds to the highest uncertainty,
and no directionality preference based on 30 posterior samples (4-6). Vectors
were projected into the UMAP space using the scVelo API based on cosine
similarity-derived transition matrix. **c**, UMAP embedded cells colored by the
average cell shared time across 30 posterior samples. The Spearman's correlation
between Pyro&thinsp;-Velocity and Cytotrace is provided. **d**,
Pyro&thinsp;-Velocity automatic gene selection and ranking for predicting
important genes involved in pancreas development. **e**, Phase portraits of
posterior samples average spliced and unspliced gene expression (left), rainbow
plots showing denoised spliced gene expression (middle), and gene expression
within UMAP embedded space.  Single dots correspond to individual cells (left,
right) or vertical lines (middle). Cell type annotation coloring is the same as
in **b**.
:::

@@

Pyro&thinsp;-Velocity accurately predicts cell fate choices and lineage
trajectory in human LARRY barcoded hematopoietic cells. Systematic benchmarking
of cell fate prediction methods - predicting uni-lineage cell fates (**a-b**, Model
1), bifurcated lineages (**c**, Model 2), or all lineages together (**d**, Model 2).
Panels from left to right show on the same SPRING embedding: 1. Cell types; 2.
Clonal progression vector field derived based on the observed clonal barcodes
([Methods](#sec-methods)); 3. scVelo recovered vector field; 4. Pyro&thinsp;-Velocity single
cell level vector field uncertainty (angular standard deviation from 30
posterior samples);
5. Pyro&thinsp;-Velocity recovered vector field; vectors are colored by the
averaged vector field uncertainty; 6. Clonal fate potency score using Cospar
[@Wang2022-xb]; 7. scVelo recovered latent time; 8. Pyro&thinsp;-Velocity
inference of posterior mean shared time per cell. Cosine similarity and
Spearman's correlation for the different methods with the clonal progression and
fate potency are reported in the titles.
:::

@@

::: {#extfig-graphical-models}
![](figures/v2/Supplementary-Figure1.pdf)

Graphical representations detail the generative process of the
Pyro&thinsp;-Velocity models and highlight their differences. **a**,
Pyro&thinsp;-Velocity Model 1. Each rectangle represents one data dimension
(e.g., gene or cell). The arrows represent the directed hierarchical dependency
among random variables. The white circles correspond to the modeled random
variables. The probability distributions of these variables are denoted on the
right legend. The gray circles correspond to intermediate results computed from
random variables. In this model, the kinetics random variables for the cell
shared time and gene are modeled with two constraints: 1. the shared time is
strictly larger than the gene-specific activation time $t_{0}^{(0)}$, 2. the
gene expression in the activate state starts with the initial condition
$(u_{0}=0,\,s_{0}=0)$, as in previous methods[@La_Manno2018-lj;@Bergen2020-pj]
([Methods](#sec-methods)). **b**, Pyro&thinsp;-Velocity Model 2. This figure follows the same
graphical convention as in **a**, In this model $t_{0}^{(0)}$ is independent from
the shared time, thus allowing a time lag for the gene activation phase. In
addition, a positive starting basal gene expression level $(u_{0},\,s_{0})$
different than $(u_{0}=0,\,s_{0}=0)$ is allowed. ([Methods](#sec-methods)). **c**, Schematic
illustrating how the constraints proposed in Model 1 (left) and Model 2 (right)
are reflected in the modeling of individual gene expression. The x-axis
corresponds to the shared time, the y-axis corresponds to the posterior mean of
the spliced expression level, the vertical lines with arrows show the start and
end time of the observed cell sampling, t0 represents the gene-specific
activation time $t_{0}^{(0)}$, the red lines highlight the differences for the
potential gene-specific time lag when $t_{0}^{(0)} > {\min}\!⁡(t)$, and the
initial spliced gene expression. The same differences can also be illustrated
for the unspliced gene expression profiles (not shown).
:::

@@

Extended version of @fig-model-results for evaluating Pyro&thinsp;-Velocity
(Model 1) and scVelo on the PBMC and pancreas datasets. **a**, Left figure shows the
scVelo vector field using the default user options, right figure shows the
evaluation of the scVelo vector field using different combinations of user
options within Epsilon cells. **b**, Pyro&thinsp;-Velocity additional single cell
level uncertainty diagnostic plots on the fully mature PBMC dataset, from left
to right: 1. standard deviation of posterior samples of shared time, 2. Rayleigh
test of posterior samples of the vector field (the title reports false discovery
rate after p-value correction for multiple tests across cells), 3.
transcriptional state uncertainty of predicted raw read count per cell. The
contour lines show the top 10% of the most uncertain cells. **c**, For the pancreas
dataset, the figures show from left to right: UMAP rendering of 1. Cytotrace, 2.
scVelo latent time, 3. Posterior mean of Pyro&thinsp;-Velocity shared time
averaged across 30 posterior samples, 4. Top 50 markers overlapping between
Pyro&thinsp;-Velocity and scVelo ([Methods](#sec-methods), Supplementary Table 1-2), 5. The same
uncertainty diagnostic plots per cell as in **b**. **d**, The same plot as [@fig-model-results]d, e.
Selection of the top negatively correlated genes with the cell shared time based
on the posterior mean on denoised spliced expression and negative mean absolute
error of gene expression prediction ([Methods](#sec-methods)). **f**, Phase portraits, rainbow
plots, and UMAP rendering of posterior mean of gene expression for the genes
selected in **e**.
:::

@@

Extension of @fig-lineage-tracing, Pyro&thinsp;-Velocity (Model 1 for uni-fate
data, Model 2 for bi-fate and multi-fate data) outperforms Cytotrace on
scRNA-seq datasets with lineage barcoding information (LARRY dataset). **a**, SPRING
embedding rendering of Cytotrace scores, the title reports the Spearman's
correlation between the Cytotrace and Cospar fate potency scores, **b**,
Pyro&thinsp;-Velocity additional single cell level uncertainty diagnostic plots
based on the multi-fate LARRY dataset, from left to right: 1. Standard deviation
of posterior samples of shared time per cell, 2. Rayleigh test of posterior
samples of vector field per cell (the title reports the false discovery rate
after p-value correction for multiple tests across cells), 3. Transcriptional
state uncertainty of predicted raw read count per cell. The contour lines show
the top 10% of the most uncertain cells ([Methods](#sec-methods)), **c**, Pairwise Spearman's
correlation between Cytotrace/Cospar fate potency scores and velocity methods
for the four different benchmarking datasets selected based on the lineage
barcoding information from LARRY data, **d**, SPRING embedding rendering of
posterior mean of shared time across 30 posterior samples, the title shows the
Spearman's correlation of mean shared time with Cytotrace. **e**, Selection of the
top positively correlated genes with the cell shared time based on the posterior
mean of spliced expression. **f**, Phase portraits, rainbow plots, and UMAP
rendering with splicing gene expression level for the selected genes in **e**.
:::

@@

Extension of [@fig-lineage-tracing] presenting an extended comparison of
Pyro&thinsp;-Velocity and scVelo on scRNA-seq datasets with lineage barcoding
information (LARRY dataset). **a**, uni-fate Monocyte lineage analysis with
Pyro&thinsp;-Velocity Model 2, **b**, uni-fate Neutrophil lineage analysis with
Pyro&thinsp;-Velocity Model 2, **c**, Neutrophil and Monocyte bi-fate lineage
analysis with Pyro&thinsp;-Velocity Model 1, and **d**, multi-fate lineage analysis
with Pyro&thinsp;-Velocity Model 1. The panel descriptions are the same as in
[@fig-lineage-tracing].
:::
